Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-10-17 papers/termination/termination-esop/paper.tex
removed aspects related to kalashnikov from the experiments

%% \subsection{Candidate Generation Strategies}
%% \noindent{\bf Explicit Proof Search.} The simplest strategy for finding candidates
%% is to just exhaustively enumerate them all, starting with the shortest and
%% progressively increasing the number of instructions.  This strategy
%% is implemented by the {\sc ExplicitSearch} routine.  Since the set of
%% $\mathcal{L}$-programs is recursively enumerable, this procedure is complete.
%% \\

%% \noindent{\bf Symbolic Bounded Model Checking.} Another complete method for generating
%% candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
%% search, we must progressively increase the length of the $\mathcal{L}$-program we search for
%% in order to get a complete search procedure.
%% \\

%% \noindent {\bf Genetic Programming and Incremental Evolution.} \label{sec:gp}
%% Our final strategy is genetic programming
%% (GP)~\cite{langdon:fogp,brameier2007linear}.  GP provides an adaptive way of
%% searching through the space of $\mathcal{L}$-programs for an individual
%% that is ``fit'' in some sense.  We measure the
%% fitness of an individual by counting the number of tests in {\sc inputs}
%% for which it satisfies the specification.

%% To bootstrap GP in the first iteration of the CEGIS loop, we generate a population
%% of random $\mathcal{L}$-programs. We then iteratively evolve this population by
%% applying the genetic operators {\sc crossover} and {\sc mutate}.
%% {\sc Crossover} combines selected existing programs into new programs,
%% whereas {\sc mutate} randomly changes parts of a single program.
%% Fitter programs are more likely to be selected.

%% % The {\sc crossover} operator takes two programs and combines their
%% % code in some way to create a new program.  Programs are selected to be
%% % {\sc crossover}ed according to their fitness -- fitter programs are more
%% % likely to be selected to become parents.

%% % The {\sc mutate} operator takes a single program and randomly changes parts
%% % of its code, again creating a new program.

%% Rather than generating a random population at the beginning of each subsequent
%% iteration of the CEGIS loop, we start with the population we had at the end of the
%% previous iteration.  The intuition here is that this population contained
%% many individuals that performed well on the $k$ inputs we had before, so
%% they will probably continue to perform well on the $k+1$ inputs we have now.
%% In the parlance of evolutionary programming, this is known as
%% incremental evolution~\cite{Gomez97incrementalevolution}.

%% % In the setting of program synthesis, genetic programming can be very
%% % usefully combined with the concept of incremental evolution.  The idea is
%% % that in order to evolve a program that is adequate for some difficult task,
%% % it is often helpful to first evolve a population that as an ensemble is good
%% % at some simpler variation of the task.  This principle leads us to the
%% % observation that each round of the CEGIS loop can be thought of as creating
%% % a more difficult problem for the GP synthesiser to solve.  Therefore, if GP
%% % finds a candidate program which is correct for the $k$ inputs so far
%% % observed, but is not correct for the full specification, that candidate
%% % program came from a population that was ``good'' for the specification
%% % restrict to those $k$ inputs.  The next time round the CEGIS loop, the
%% % problem GP must solve is slightly harder -- any solution to this new problem
%% % must be correct for the previous $k$ inputs and also whatever new input was
%% % added by the previous iteration of {\sc verif}.  It therefore makes sense to
%% % begin the $k+1$st round of evolution with the population at the end of the
%% % $k$-th round.

\section{Soundness, Completeness and Complexity}
The second-order SAT solver we are using is a semi-decision procedure -- for any
satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables~\cite{kalashnikov}. 
%% Termination for bit-vector programs is known to be
%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
%% expected results for soundness and completeness of our algorithm for
%% deciding it.  Our strategy for proving soundness and completeness
%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
%% with \newC as a background logic) is a semi-decision
%% procedure.
In this section, we show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our technique terminates with a valid proof for every input program.
%
%% \begin{theorem}\label{thm:synth-sound}
%% Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $(P, x_0)$, then
%% $(P, x_0) \models \sigma$.
%% \end{theorem}

%% \begin{proof}
%%  The procedure {\sc Synthesise} terminates only if {\sc Verif} returns ``valid''.  In that
%%  case, $\exists x . \lnot \sigma(P, x_0, x)$ is unsatisfiable and so $\forall x . \sigma(P, x_0, x)$ holds.
%% \end{proof}

%% \begin{theorem}
%%  \label{thm:synth-semicomplete}
%%  If the existential first-order theory used to express the specification $\sigma$ is decidable
%%  and the domain of inputs $X$ is finite,
%%  Algorithm~\ref{alg:cegis} is semi-complete -- if a solution $(P, x_0) \models \sigma$
%%  exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
%%  satisfies the specification, the algorithm may not terminate.
%% \end{theorem}

%% \begin{proof}
%%  If the domain $X$ is finite then the loop in procedure {\sc Synthesise} can only
%%  iterate $\| X \|$ times, since by this time all of the elements of $X$ would have been
%%  added to the inputs set.  Therefore if the {\sc Synth} procedure always terminates,
%%  Algorithm~\ref{alg:cegis} does as well.

%%  Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
%%  the program length $l$), it will eventually enumerate a program that meets the specification
%%  on whatever set of inputs are currently being tracked, since by assumption such a program
%%  exists.  Since the first-order theory is
%%  decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
%%  The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
%%  enumerates this set, so it is semi-complete.
%% \end{proof}

%% \begin{corollary}
%% \label{cor:cegis-newc}
%%  Since safety of \newC programs is decidable, Algorithm~\ref{alg:cegis} is semi-complete
%%  when instantiated with \newC as a background theory.
%% \end{corollary}

%% % \begin{theorem}
%% % \label{thm:sound}
%% %  {\sc Headshot} is sound -- if it returns a verdict of termination, the program under
%% %  analysis terminates on all inputs.  Similarly if it returns a verdict of non-termination,
%% %  the program under analysis has at least one infinite execution.
%% % \end{theorem}
%% % 
%% % \begin{proof}
%% %  This is a direct consequence of Theorems~\ref{thm:ut}, \ref{thm:ct}, \ref{thm:ont}, \ref{thm:cnt} and \ref{thm:synth-sound}.
%% % \end{proof}

%% We will now show that $\mathcal{L}$ is expressive enough to encode
%% \mbox{(non-)}ter\-mi\-nation proofs for every bit-vector program.

@@

%%---- Hence, Algorithm~\ref{alg:termination} is sound and complete for bit-vector termination.

%% Theorem~\ref{thm:synth-semicomplete} provides a time bound -- the CEGIS loop can iterate at most
%% $2^{bk}$ times where $k$ is the number of variables used by $L$ and
%% $b$ is the number of bits used to store each variable.
%% The $i$th iteration of the loop requires solving 2 NP-complete problems of size $i \times l \times \| L\|$
%% with $l$ the length of the program to be synthesised, and $\| L \| $ the size of the program-under-analysis
%% computing $ L $.  So if
%% the CEGIS loop iterates $n$ times, the
%% runtime is bounded by
%% $$2 \times n \times NP(n \times l \times \| L \|) \leq 2^{bk + 1} NP(2^{bk} \times l \times \| L \|)$$
%% Clearly the runtime is affected most strongly by $n$, the total number of iterations of the CEGIS loop.

%% We will now show that the number of iterations of the CEGIS loop is a function
%% of the Kolmogorov complexity of the \mbox{(non-)}ter\-mi\-nation proof, and so
%% our runtime is asymptotically a function of the Kolmogorov complexity.
%% We argue that this gives our procedure various desirable qualities in practical
%% applications.
%% We first recall the definition of the Kolmogorov complexity
%% of a function $f$:

%% \begin{definition}[Kolmogorov complexity]
%%  The Kolmogorov complexity $K(f)$ is the length of the shortest program that
%%  computes~$f$.
%% \end{definition}

%% We can extend this definition slightly to talk about the Kolmogorov complexity of a
%% synthesis problem in terms of its specification:

%% \begin{definition}[Kolmogorov complexity of a synthesis problem]
%%  The Kolmogorov complexity of a program specification $K(\sigma)$ is the length of the shortest
%%  program $P$ such that $P \models \sigma$.
%% \end{definition}

%% Let us consider the number of iterations of the CEGIS loop $n$ required for a specification
%% $\sigma$.  Since we enumerate candidate programs in order of length, we are always synthesising
%% programs with length no greater than $K(\sigma)$ (since when we enumerate the first correct program,
%% we will terminate).  So the space of solutions we search over is the space
%% of functions computed by $\mathcal{L}$-programs of length no greater than $K(\sigma)$.  Let's
%% denote this set $\mathcal{L}(K(\sigma))$.
%% Since there are $O(2^{K(\sigma)})$ \emph{programs} of length $K(\sigma)$ and some functions
%% will be computed by more than one program, we have $\| \mathcal{L}(K(\sigma)) \| \leq O(2^{K(\sigma)})$.

%% Each iteration of the CEGIS loop distinguishes at least one incorrect function from the set of correct
%% functions, so the loop will iterate no more than $\| \mathcal{L}(K(\sigma)) \|$ times.
%% Therefore another bound on our runtime is:
%% $$O\left (2^{K(\sigma)} \times NP \left ( K(\sigma) \times \| L \| \right) \right) = O \left(2^{K(\sigma)} \right)$$
%% So for even moderately large $K(\sigma)$ our procedure is intractably slow.  Conversely
%% if $K(\sigma)$ is small we will be able to find a termination proof, regardless of other parameters
%% such as the size of the program under analysis.  To put this another way, under the assumption
%% that $K(\sigma)$ is small for a program's termination problem, we will find a termination proof
%% quite rapidly.

%% To give a flavour for the nature of the ``low Kolmogorov complexity'' assumption, 
%% we observe that linear functions have low Kolmogorov complexity (since a linear function can
%% be computed with at most 1 instruction and 1 multiplication for each program variable),
%% so the number of linear functions $f$ with $K(f) \leq \kappa$ grows linearly with $\kappa$.
%% Conversely, the number of functions $f$ with $K(f) \leq \kappa$
%% grows exponentially with $\kappa$.  This shows that assuming the existence of a linear
%% ranking function is much stronger than assuming the existence of a short
%% termination proof.  In addition to being a much weaker assumption,
%% we argue that low Kolmogorov complexity is a more natural assumption than
%% linearity.  Humans tend to write programs they can understand.  It's hard
%% to quantify exactly what ``understandable'' means, but we think that
%% ``having a short proof'' is closer to the mark than ``having a linear proof''.

@@

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover,
named {\sc Juggernaut}, on 47
benchmarks taken from the literature and
SV-COMP'15~\cite{svcomp15}\footnote{The logs and benchmarks are included in the
non-anonymous supplementary material submitted with this paper}.%\footnote{The benchmarks, code and log files may be downloaded from the URL \texttt{http://www.cprover.org/juggernaut/}}.
To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of 180\,s, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

@@

The final column in Figure~\ref{fig:experiments} gives the number of
iterations of the CEGIS loop our synthesiser performed.  While the
theoretical bound for the number of CEGIS iterations is very large
(exponential in the amount of space used by the program), these results
demonstrate that in practice the number of iterations required is usually
small.

Of the 47 benchmarks, 2 use nonlinear operations in the program (loop6 and loop11),
and 5 have nested loops (svcomp6, svcomp12, svcomp18, svcomp40, svcomp41).
{\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
It solves 4 of the 5 nested loops in less than 30\,s\footnote{Although
2 of these cases timed out with our default parameters, changing one command
line flag allowed them to go through.  The benchmarks using non-default
command line flags are marked with a *.}, but times out on the 5th.

Overall, we notice that the GP component was always the fastest to find
candidate solutions, {\sc ExplicitSearch} was always fastest
to find counterexample inputs, and {\sc CBMC} was always fastest at
proving correctness.  Each of these search methods has its strengths,
and removing any one would have had a significant negative impact on
overall performance.  Approximately 90\,\% of the total
runtime was spent in synthesising new candidate solutions, which matches
our expectations -- the (non-)termination proofs are difficult to
find, but easy to check.

In conclusion,
these experiments confirm our conjecture that synthesis can be used
effectively to prove termination and non-termination.  In particular,
for programs with nested loops, nonlinear arithmetic and complex
termination arguments, the versatility of unrestricted synthesis
is very valuable.

\begin{figure}
\centering
\scriptsize
\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|r|}
\hline
          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{3}{|c|}{\sc Juggernaut} \\
Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time & \#It.\\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}

@@

Times marked with a * were run with non-default command line flags.

>>>>

\section{Soundness, Completeness and Complexity}
The second-order SAT solver we are using is a semi-decision procedure -- for any
satisfiable formula, the solver is guaranteed to find a satisfying assignment to
all the second-order variables~\cite{kalashnikov}. 
%% Termination for bit-vector programs is known to be
%% PSPACE-complete~\cite{DBLP:conf/tacas/CookKRW10}.  We first record the
%% expected results for soundness and completeness of our algorithm for
%% deciding it.  Our strategy for proving soundness and completeness
%% is first to prove that Algorithm~\ref{alg:cegis} (as instantiated
%% with \newC as a background logic) is a semi-decision
%% procedure.
In this section, we show that $\mathcal{L}$ is expressive enough
to compute (non-)termination proofs for every bit-vector program.
By using this result, %We will then combine this result with the fact that the second-order SAT solver is a semi-decision procedure to 
we then show that our technique terminates with a valid proof for every input program.

@@

@@

To evaluate our algorithm, we implemented a tool that generates a
termination specification from a C program and calls the second-order SAT solver in \cite{kalashnikov} to obtain a proof.
We ran the resulting termination prover,
named {\sc Juggernaut}, on 47
benchmarks taken from the literature and
SV-COMP'15~\cite{svcomp15}.
To provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of 180\,s, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

@@

%% The final column in Figure~\ref{fig:experiments} gives the number of
%% iterations of the CEGIS loop our synthesiser performed.  While the
%% theoretical bound for the number of CEGIS iterations is very large
%% (exponential in the amount of space used by the program), these results
%% demonstrate that in practice the number of iterations required is usually
%% small.

Of the 47 benchmarks, 2 use nonlinear operations in the program (loop6 and loop11),
and 5 have nested loops (svcomp6, svcomp12, svcomp18, svcomp40, svcomp41).
{\sc Juggernaut} handles the nonlinear cases correctly and rapidly.
It solves 4 of the 5 nested loops in less than 30\,s, but times out on the 5th.

%% Overall, we notice that the GP component was always the fastest to find
%% candidate solutions, {\sc ExplicitSearch} was always fastest
%% to find counterexample inputs, and {\sc CBMC} was always fastest at
%% proving correctness.  Each of these search methods has its strengths,
%% and removing any one would have had a significant negative impact on
%% overall performance.  Approximately 90\,\% of the total
%% runtime was spent in synthesising new candidate solutions, which matches
%% our expectations -- the (non-)termination proofs are difficult to
%% find, but easy to check.

In conclusion,
these experiments confirm our conjecture that second-order SAT can be used
effectively to prove termination and non-termination.  In particular,
for programs with nested loops, nonlinear arithmetic and complex
termination arguments, the versatility given by a general purpose solver %unrestricted termination arguments %synthesis
is very valuable.

\begin{figure}
\centering
\small
\begin{tabular}{|l|@{}c@{}||@{}c@{}|r||@{}c@{}|r|}
\hline
          &             & \multicolumn{2}{|c||}{\sc ARMC} & \multicolumn{2}{|c|}{\sc Juggernaut} \\
Benchmark & \,Expected\, & \,Verdict\, & Time & \,Verdict\, & Time \\
    \hline
    \hline
\input{table}
    \hline
\end{tabular}

@@
